The effect of taxonomic bias on differential-abundance analysis in microbiome studies

### 2022-07-10 case-studies.Rmd
Change 'sign error' to 'direction error'

We sought to understand the potential impact of bias on DA analysis of vaginal microbiome profiles from the MOMS-PI study.
We first examined variation in the mean efficiency across samples under the assumption that bias in the MOMS-PI study was accurately represented by the @brooks2015thet control measurements.
Using taxonomic relatedness to impute the efficiencies of species not in the controls, we were able to calibrate the MOMS-PI profiles and estimate the mean efficiency of each sample (REF SI Analysis).
The mean efficiency varies substantially across vaginal samples (Figure 4.1), with samples in which a _Lactobacillus_ species is most abundant typically having a mean efficiency that is 3-fold to 20-fold greater than samples in which _G. vaginalis_ is most abundant.
The vaginal microbiome sometimes shifted between _Lactobacillus_-dominance and _Gardnerella_-dominance between consecutive visits in individual women (SI Figure \@ref(fig:momspi-mean-efficiency-fcs)), which caused magnitude and sign errors in the trajectories of lower-abundance species (SI Figure \@ref(fig:momspi-trajectory)).

>>>>

We sought to understand the potential impact of bias on DA analysis of vaginal microbiome profiles from the MOMS-PI study.
We first examined variation in the mean efficiency across samples under the assumption that bias in the MOMS-PI study was accurately represented by the @brooks2015thet control measurements.
Using taxonomic relatedness to impute the efficiencies of species not in the controls, we were able to calibrate the MOMS-PI profiles and estimate the mean efficiency of each sample (REF SI Analysis).
The mean efficiency varies substantially across vaginal samples (Figure 4.1), with samples in which a _Lactobacillus_ species is most abundant typically having a mean efficiency that is 3-fold to 20-fold greater than samples in which _G. vaginalis_ is most abundant.
The vaginal microbiome sometimes shifted between _Lactobacillus_-dominance and _Gardnerella_-dominance between consecutive visits in individual women (SI Figure \@ref(fig:momspi-mean-efficiency-fcs)), which caused magnitude and direction errors in the trajectories of lower-abundance species (SI Figure \@ref(fig:momspi-trajectory)).

### 2022-07-10 supplemental-figures.Rmd
Change 'sign error' to 'direction error'

(ref:cap-momspi-trajectory) **In vaginal microbiome measurements, shifts between _Lactobacillus_ and _Gardnerella_ dominance can drive spurious fold changes in other, lower-abundance species.** The figure shows species proportions and mean efficiency trajectories over consecutive clinical visits for a subject in the MOMS-PI study whose microbiome samples showed substantial variation in mean efficiency. The subject's samples are dominated by _Gardnerella vaginalis_ and _Lachnospiraceae BVAB1_ during the first three visits before transitioning to being dominated by _Lactobacillus iners_ between visits 3 and 4. This transition drives a sharp increase in the mean efficiency, which significantly distorts the fold changes in the observed (uncalibrated) microbiome measurements for species with less dramatic fold changes. Two exemplar species are shown to illustrate the magnitude (_Ureaplasma cluster 23_) and sign (_Megasphaera OTU70 type1_) errors that can arise in this situation.

>>>>

(ref:cap-momspi-trajectory) **In vaginal microbiome measurements, shifts between _Lactobacillus_ and _Gardnerella_ dominance can drive spurious fold changes in other, lower-abundance species.** The figure shows species proportions and mean efficiency trajectories over consecutive clinical visits for a subject in the MOMS-PI study whose microbiome samples showed substantial variation in mean efficiency. The subject's samples are dominated by _Gardnerella vaginalis_ and _Lachnospiraceae BVAB1_ during the first three visits before transitioning to being dominated by _Lactobacillus iners_ between visits 3 and 4. This transition drives a sharp increase in the mean efficiency, which significantly distorts the fold changes in the observed (uncalibrated) microbiome measurements for species with less dramatic fold changes. Two exemplar species are shown to illustrate the magnitude (_Ureaplasma cluster 23_) and direction (_Megasphaera OTU70 type1_) errors that can arise in this situation.
